Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-24 paper.qmd
revision of intro

A more robust approach than simply listing package versions is to use the
[renv]{.pkg} package. [renv]{.pkg} captures the project's software state and
writes it to a `renv.lock` file, which includes the exact versions of
[R]{.proglang} and all required packages. This lockfile serves as a blueprint
for restoring the environment automatically, ensuring that others can recreate
the same setup with minimal effort. Below is an abbreviated example of an
`renv.lock` file:
```{=tex}
\begin{Code}
{
  "R": {
    "Version": "4.2.2",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://packagemanager.rstudio.com/all/latest"
      }
    ]
  },
  "Packages": {
    "MASS": {
      "Package": "MASS",
      "Version": "7.3-58.1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "762e1804143a332333c054759f89a706",
      "Requirements": []
    },
    "Matrix": {
      "Package": "Matrix",
      "Version": "1.5-1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "539dc0c0c05636812f1080f473d2c177",
      "Requirements": ["lattice"]
    }
    ... additional packages omitted ...
  }
}
\end{Code}
```

@@

However, [renv]{.pkg} has important limitations. It does not restore the version
of [R]{.proglang} itself—installing the correct R version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg}
cannot resolve this conflict. Users must manually install system libraries, and
the process differs across operating systems. Further details are discussed in
the [renv]{.pkg} documentation^[See
\url{https://rstudio.github.io/renv/articles/renv.html}.].

Other packages provide similar functionality with different trade-offs.
[groundhog]{.pkg} by @simonsohn2023 makes it easy to install packages as they
existed on CRAN at a given date:

::: {.content-hidden when-format="pdf"}
```R
groundhog.library("
 library('purrr')
 library('ggplot2')",
 "2017-10-04",
 tolerate.R.version = "4.2.2")
```
:::

```{=tex}
\begin{CodeInput}
R> groundhog.library("
+   library('purrr')
+   library('ggplot2')",
+   "2017-10-04",
+   tolerate.R.version = "4.2.2")
\end{CodeInput}
```

[groundhog]{.pkg} places packages in project-specific libraries but does not
install [R]{.proglang} itself. Without the `tolerate.R.version` argument, it
will refuse to proceed if the R version does not match expectations.
[rang]{.pkg}, developed by @chan2023, offers similar date-based package
installation. The Posit Package Manager provides another approach through dated
CRAN snapshots, configured via the `.Rprofile` file:

::: {.content-hidden when-format="pdf"}
```R
options(repos =
  c(REPO_NAME =
    "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
  )
)
```
:::

```{=tex}
\begin{CodeInput}
R> options(repos =
+    c(REPO_NAME =
+      "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
+    )
+  )
\end{CodeInput}
```

This approach installs all packages from the specified snapshot date, but unless
users explicitly manage separate libraries for different projects, all projects
will share the same package versions.

These tools represent significant progress in reproducibility, but they share a
common limitation: none addresses system-level dependencies or supports polyglot
environments. A project using [R]{.proglang}, [Python]{.proglang}
[@python3], and system tools like [Quarto]{.pkg} or \LaTeX requires coordinating
multiple package managers ([renv]{.pkg} for R, virtual environments for
[Python]{.proglang}, manual installation for system tools), each with its own
configuration and potential for conflicts.

@@

4. Single-language focus: While [Docker]{.pkg} can support polyglot
  environments, orchestrating them requires manual coordination. There is no
  standard way to specify that one step should run in a [Python]{.pkg}
  environment while another runs in R.

5. Incomplete Reproducibility: Reproducibility should be viewed as a spectrum
   rather than a binary property, as emphasised in *Reproducibility in Software
   Engineering* by @dellaieraMasterThesis2024. [Docker]{.pkg} illustrates this
   spectrum well: it achieves strong run-time reproducibility—consistency across
   systems (space)—but weak build-time reproducibility, which concerns
   consistency across time. As defined in the thesis, build-time reproducibility
   means producing identical artefacts regardless of when or where they are
   built. [Docker]{.pkg} cannot guarantee this, as its reliance on mutable base
   images and non-deterministic commands such as `apt-get update` (or similar
   commands, depending on which base image is used) introduces temporal drift:
   rebuilding the same `Dockerfile` at different times can yield different
   results. Even with pinned base images, the thesis found that container
   checksums varied across builds and machines. Thus, while [Docker]{.pkg}
   provides partial reproducibility, strong in space but weak in time, verifiable,
   deterministic artefacts require pinning image digests, freezing dependencies,
   and using deterministic build systems such as [Nix]{.pkg} for full control
   over both dimensions [@Malka_2024].

>>>>

A more robust approach than simply listing package versions is to use tools that
help data scientists define per-project environments. This is common practice
for [Python]{.proglang} users, who have many tools at their disposal, such as
the built-in *virtual environments* module, but also sophisticated third-party
tools. One particularly popular tool is [uv]{.pkg}, which manages both
project-specific [Python]{.proglang} interpreters (enabling different Python
versions per project) and package libraries using a lock file mechanism.

In the [R]{.proglang} ecosystem, virtual environments are not a built-in
language feature as they are in [Python]{.proglang}, but tools like [renv]{.pkg}
provide equivalent functionality. [renv]{.pkg} captures the project's software
state and writes it to a lock file (called `renv.lock`), which includes the
exact versions of [R]{.proglang} and all required packages. This lockfile serves
as a blueprint for restoring the environment automatically, ensuring that others
can recreate the same setup with minimal effort. Below is an abbreviated example
of an `renv.lock` file:

```{=tex}
\begin{Code}
{
  "R": {
    "Version": "4.2.2",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://packagemanager.rstudio.com/all/latest"
      }
    ]
  },
  "Packages": {
    "MASS": {
      "Package": "MASS",
      "Version": "7.3-58.1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "762e1804143a332333c054759f89a706",
      "Requirements": []
    },
    "Matrix": {
      "Package": "Matrix",
      "Version": "1.5-1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "539dc0c0c05636812f1080f473d2c177",
      "Requirements": ["lattice"]
    }
    ... additional packages omitted ...
  }
}
\end{Code}
```

@@

However, both [renv]{.pkg} and [uv]{.pkg} have important limitations. While
[uv]{.pkg} manages Python interpreter versions directly, [renv]{.pkg} does not
restore the version of [R]{.proglang} itself—installing the correct R version
must be done separately using tools like [rig]{.pkg} [@rlib2023]. More
critically, neither [uv]{.pkg} nor [renv]{.pkg} handle system-level
dependencies. If [sf]{.pkg} requires [GDAL]{.pkg} version 3.0 but the system has
version 2.4 installed, neither tool can resolve this conflict. Users must
manually install system libraries, and the process differs across operating
systems. Further details are discussed in the [renv]{.pkg} documentation^[See
https://rstudio.github.io/renv/articles/renv.html.].

Pre-compiled binary packages (Python wheels or R binaries from Posit Package
Manager) can simplify installation by avoiding compilation, but they do not
eliminate the need for system libraries that would be required at runtime—these
must still be installed separately.

Other [R]{.proglang} packages provide similar functionality with different
trade-offs. [groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023]
enable date-based package installation, retrieving packages as they existed on
CRAN at a specified date. The Posit Package Manager offers another approach
through dated CRAN snapshots. However, like [renv]{.pkg}, these tools focus on
[R]{.proglang} packages and do not address system-level dependencies or
[R]{.proglang} version management.

These tools represent significant progress in reproducibility within their
respective ecosystems, but they share a common limitation: none addresses
system-level dependencies or supports polyglot environments. A project using
[R]{.proglang}, [Python]{.proglang} [@python3], and system tools like
[Quarto]{.pkg} or \LaTeX requires coordinating multiple package managers
([renv]{.pkg} for R, [uv]{.pkg} for [Python]{.proglang}, manual installation for
system tools), each with its own configuration and potential for conflicts.

@@

4. Single-language focus: While [Docker]{.pkg} can support polyglot
  environments, orchestrating them requires manual coordination. There is no
  standard way to specify that one step should run in a [Python]{.proglang}
  environment while another runs in R.

5. Incomplete Reproducibility: Reproducibility should be viewed as a spectrum
   rather than a binary property, as emphasised in *Reproducibility in Software
   Engineering* by @dellaieraMasterThesis2024. [Docker]{.pkg} illustrates this
   spectrum well: it achieves strong run-time reproducibility—consistency across
   systems (space)—but weak build-time reproducibility, which concerns
   consistency across time. As defined in the thesis, build-time reproducibility
   means producing identical artefacts regardless of when or where they are
   built. [Docker]{.pkg} cannot guarantee this, as its reliance on mutable base
   images and non-deterministic commands such as `apt-get update` (or similar
   commands, depending on which base image is used) introduces temporal drift:
   rebuilding the same `Dockerfile` at different times can yield different
   results. Even with pinned base images, the thesis found that container
   checksums varied across builds and machines. Thus, while [Docker]{.pkg}
   provides strong run-time reproducibility, achieving verifiable, deterministic
   artefacts requires pinning image digests, freezing dependencies, and using
   deterministic build systems such as [Nix]{.pkg} for full control over both
   temporal and spatial dimensions [@Malka_2024].
